### 2022-08-09 main/chapter_2ljets.tex
add last of discussion. autocheck

\begin{figure}[tp]
\centering
\includegraphics[width=0.48\textwidth]{figures/2ljets_presel_met_logy.png}
\hfill
\includegraphics[width=0.48\textwidth]{figures/2ljets_presel_met_sig_logy.png}
\caption[
Illustration of how $\metsig$ beats $\met$ in sensitivity to example signal
models
]{
Illustration of how $\metsig$ beats $\met$ in sensitivity to example signal
models.
At large $\metsig$, $\ttbar$ and other backgrounds are suppressed, leaving
quite pure \diboson\ backgrounds and greater significance measures.
Signal models with smaller mass-splittings tend to have the bulk of their
events at smaller $\met$.
\\[0.5em]
This selection requires two same-flavour opposite-sign signal leptons with
$\pt^{\ell_2} > 25\,\eV[G]$,
$\mll \in (71, 111)\,\eV[G]$,
$\mjj \in (60, 110)\,\eV[G]$,
$\nbtag \leq 1$,
$\met > 100\,\eV[G]$, and
$\metsig > 6$.
\\[0.5em]
``Significance'' displayed in the lower subplot uses the Binomial significance
$Z_{Bi}$ from~\cite{cousins2008evaluation}, with a $30\%$ background
uncertainty and yields taken to the right of a given value.
Large values indicate expected sensitivity to the signal model.
}
\label{fig:2ljets_presel_met}
\end{figure}

@@

Although my style of spelling out FNP or FNP every time
(as opposed to just `fake') may be awkward, it is supported by the fact that this
the non-prompt part dominates.
Indeed, figures in our internal documentation~\cite{twoljets2018int} suggest
that, in simulations, non-prompt hadron decays produce in excess of $90\%$ of
this background in a loose selection near to our regions, and across the
entire $\metsig$ spectrum.
Non-prompt muons are mostly from decays of heavy-flavour ($b$ or $c$) hadrons,
and electrons arise in roughly equal parts from light flavour hadrons.
Tau decays form the next largest part, with prompt fakes due to misidentified
photons or leptons in the sub-$\%$ weeds.

@@

To validate FNP yields in a region with sufficiently many data for precision,
additional validations were also inspected with different-flavour or
same-charge requirements, and all showed reasonably agreement with data
within uncertainties.
In particular, a region named VR-SS is defined to inspect data with
same-charged (same sign, SS) leptons in a loose preselection that covers all
of our main regions;%
\footnote{%
Like many of the regions plotted in designing this search, VR-SS is not a
`core' validation region in the sense that it is not reported in the paper.
Perhaps a fourth categorization would be useful for such regions to supplement
CR, VR, and SR, but it should be clear that any VR not included in the main
region definitions is not core and not reported in the paper.
}
VR-SS requires the pre-selections of Table~\ref{tab:2ljets_presel},
but replaces its requirement of opposite-charge leptons with a requirement of
same-charge leptons, and adds a requirement of $\mll \in (12, 111)~\eV[G]$
to cover both on- and off-shell selections.

@@

Jet mass resolution uncertainties were evaluated by smearing simulations
to increase their resolution error (decrease their precision) by $20\%$
and comparing the smeared results to standard results according to a standard
prescription~\cite{atlas_twiki_jmr}.
These apply equally to \afii\ and FullSim samples.

\emph{This jet mass modelling was discussed with, and approved by, the \atlas\
Jet/Etmiss conveners. I owe special thanks to Chris~Delitzsch for assistance
with the official jet mass scale variations.}

@@

\begin{figure}[tp]
\centering
\includegraphics[width=0.6\textwidth]{figures/2ljets_drhigh_plot_n0.pdf}
\caption[
Test statistic distribution for DR-High close to 3
]{%
Test statistic distribution for DR-High close to $3$.
This zooms in to the narrow spike of toy samples that should have $n=0$
observed data in the discovery region.
Because the likelihood function for zero data,
$\mathrm{Poisson}(0\mid x) = e^{-x}$, is $\log$-linear,
this would analytically be a delta spike.
It acquires this width from the finite precision of numerical optimization
algorithms.
The observed test statistic falling far to the right of that numerical noise
reduces our statistical precision in estimating $\clb$ and $\clsb$.
}
\label{fig:2ljets_test_stat_distributions_drhigh_n0}
\end{figure}

Although I have found no \atlas\ search reporting a discovery region limit
below $3$, there are published examples with zero data but larger observed
limits~\cite{SUSY-2018-36, atlas_susy_strong_0l}.
The most extreme is $\mathrm{SR}3\ell_{[90,110]}$
of~\cite{SUSY-2018-36}, which reports a background expectation of
$1.08 \pm 0.18$ and expected limits of
$S_{\mathrm{exp}}^{95} = 3.8^{+1.3}_{-0.5}$, but zero observed data with a
whopping $S_{\mathrm{obs}}^{95} = 4.8$.
Using their \pyhf\ models published through \hepdata\ I have failed to
reproduce this effect outside the \histfitter-\roofit ecosystem, but would
not be surprised if similar complications with the test statistic distribution
played a part.

@@

These searches do not exist in isolation. They extend and strengthen the
results of numerous previous \atlas\ searches that have used similar or smaller
datasets and found similar conclusions.
By providing refined and updated interpretations, we have advanced \atlas' SUSY
research programme in this $2\ell$+jets+$\met$ channel.
My focus has been on the $\twoljets$-electroweak search within this project.
Its result is a central pillar of the visible effects of both C1N2 and GMSB
simplified supersymmetric models, and uniquely contributes exclusion
(under the $95\%$ $\cls$ prescription) of certain regions of the models'
parameter spaces.

@@

% limitations
% overly ambitious design
% lack: resources to study, methods to analyse
% cost: delays, opportunity cost
% higher quality result by simpler more robust design published sooner
% accelerate science by doing it now, not in two years.

>>>>

\begin{figure}[tp]
\centering
\includegraphics[width=0.48\textwidth]{figures/2ljets_presel_met_logy.png}
\hfill
\includegraphics[width=0.48\textwidth]{figures/2ljets_presel_met_sig_logy.png}
\caption[
Illustration of how $\metsig$ beats $\met$ in sensitivity to example signal
models
]{
Illustration of how $\metsig$ beats $\met$ in sensitivity to example signal
models.
At large $\metsig$, $\ttbar$ and other backgrounds are suppressed, leaving
quite pure \diboson\ backgrounds and greater significance measures.
Signal models with smaller mass-splittings tend to have the bulk of their
events at smaller $\met$.
\\[0.5em]
This selection requires two same-flavour opposite-sign signal leptons with
$\pt^{\ell_2} > 25\,\eV[G]$,
$\mll \in 71\textrm{--}111\,\eV[G]$,
$\mjj \in 60\textrm{--}110\,\eV[G]$,
$\nbtag \leq 1$,
$\met > 100\,\eV[G]$, and
$\metsig > 6$.
\\[0.5em]
``Significance'' displayed in the lower subplot uses the Binomial significance
$Z_{Bi}$ from~\cite{cousins2008evaluation}, with a $30\%$ background
uncertainty and yields taken to the right of a given value.
Large values indicate expected sensitivity to the signal model.
}
\label{fig:2ljets_presel_met}
\end{figure}

@@

Although my style of spelling out FNP or FNP every time
(as opposed to just `fake') may be awkward, it is supported by the fact that this
the non-prompt part dominates.
Indeed, figures in our internal documentation~\cite{twoljets2018int} suggest
that, in simulations, non-prompt hadron decays produce in excess of $90\%$ of
this background in a loose selection near to our regions, and across the
entire $\metsig$ spectrum.
Non-prompt muons are mostly from decays of heavy-flavour ($b$ or $c$) hadrons,
and electrons arise in roughly equal parts from light flavour hadrons.
Tau decays form the next largest part, with prompt fakes due to misidentified
photons or leptons in the sub-percent weeds.

@@

To validate FNP yields in a region with sufficiently many data for precision,
additional validations were also inspected with different-flavour or
same-charge requirements, and all showed reasonably agreement with data
within uncertainties.
In particular, a region named VR-SS is defined to inspect data with
same-charged (same sign, SS) leptons in a loose preselection that covers all
of our main regions;%
\footnote{%
Like many of the regions plotted in designing this search, VR-SS is not a
`core' validation region in the sense that it is not reported in the paper.
Perhaps a fourth categorization would be useful for such regions to supplement
CR, VR, and SR, but it should be clear that any VR not included in the main
region definitions is not core and not reported in the paper.
}
VR-SS requires the pre-selections of Table~\ref{tab:2ljets_presel},
but replaces its requirement of opposite-charge leptons with a requirement of
same-charge leptons, and adds a requirement of $\mll \in 12\textrm{--}111\,\eV[G]$
to cover both on- and off-shell selections.

@@

Jet mass resolution uncertainties were evaluated by smearing simulations
to increase their resolution error (decrease their precision) by $20\%$
and comparing the smeared results to standard results according to a standard
prescription~\cite{atlas_twiki_jmr}.
These apply equally to both \afii\ and FullSim samples.

\emph{This jet mass modelling was discussed with, and approved by, the \atlas\
Jet/Etmiss conveners. Special thanks to Chris~Delitzsch for assistance
preparing and sharing the official jet mass scale variations.}

@@

\begin{figure}[tp]
\centering
\includegraphics[width=0.6\textwidth]{figures/2ljets_drhigh_plot_n0.pdf}
\caption[
Test statistic distribution for DR-High close to $3$
]{%
Test statistic distribution for DR-High close to $3$.
This zooms in to the narrow spike of toy samples that should have $n=0$
observed data in the discovery region.
Because the likelihood function for zero data,
$\mathrm{Poisson}(0\mid x) = e^{-x}$, is $\log$-linear,
this would analytically be a delta spike.
It acquires this width from the finite precision of numerical optimization
algorithms.
The observed test statistic falling far to the right of that numerical noise
reduces our statistical precision in estimating $\clb$ and $\clsb$.
}
\label{fig:2ljets_test_stat_distributions_drhigh_n0}
\end{figure}

Although I have found no \atlas\ search reporting a discovery region limit
below $3$, there are published examples with zero data but larger observed
limits~\cite{SUSY-2018-36, atlas_susy_strong_0l}.
The most extreme is $\mathrm{SR}3\ell_{[90,110]}$
of~\cite{SUSY-2018-36}, which reports a background expectation of
$1.08 \pm 0.18$ and expected limits of
$S_{\mathrm{exp}}^{95} = 3.8^{+1.3}_{-0.5}$, but zero observed data with a
whopping $S_{\mathrm{obs}}^{95} = 4.8$.
Using their \pyhf\ models published through \hepdata, I have failed to
reproduce this effect outside the \histfitter-\roofit system, finding instead
a limit close to $3$. It would not surprise me if similar complications with
the test statistic distribution played a part.

@@

These searches do not exist in isolation; they extend and strengthen the
results of numerous previous \atlas\ searches that have used similar or smaller
datasets and found similar conclusions.
By providing refined and updated interpretations, we have advanced \atlas' SUSY
research programme in this $2\ell$+jets+$\met$ channel.
My focus has been on the $\twoljets$-electroweak search within this project.
Its result is a central pillar of the visible effects of both C1N2 and GMSB
simplified supersymmetric models, and uniquely contributes exclusion
(under the $95\%$ $\cls$ prescription) of certain regions of the models'
parameter spaces.

@@

% limitations
Such successes do not, however, deny that there is room for improvement.
Although I brazenly philosophized in
Section~\ref{sec:2ljets_origins_electroweak} that the limitations of dubious
statistical practices should not prevent us from using large numbers of small
signal regions, their overly ambitious design in the $\twoljets$-electroweak
did come with costs:
% mm fakes
in replacing unusable fake/non-prompt assignments
(Section~\ref{sec:2ljets_mm_fakes}),
% precise alternative simulations
in failures of the statistical precision of alternative simulated samples
(Section~\ref{sec:2ljets_sample_comparisons})
and struggles to reconcile their standard prescriptions for coupled
(correlated) variations with physical intuition between our many bins,
% computational
and computational costs particularly in the `toy' simulations that assign our
statistical results (Section~\ref{sec:2ljets_results}).
The root of this problem that in diverging from standard designs we
had to use tools and rules outside the domains that their designers
intended.
No design can be expected to work perfectly outside its expected context,
and this divergence incurred us costs in either engineering workarounds or
leaving dubious features in the final result.

This thesis has supplemented many details of such features that are not
explicitly declared in the main publication~\cite{atlas2022searches},
and might otherwise have remained hidden in \atlas' internal documentation.

Perhaps we \emph{should not} be restrained by crude statistical
methods, but the fact is that we \emph{are} limited to immutable standard
prescriptions.
And by ignoring that truth, my designs cost this search project complications
that caused real delays when problematic data
--- substantial deficits ---
were observed in our complicated and rigidly-modelled signal regions.
Although these regions are sensitive, as demonstrated by the extent of their
projected and observed exclusion contours, delays are themselves a meaningful
cost.
Simpler designs, with fewer and larger regions, could be cheaper to model with
our limited simulation capacity and more robust to unexpected data, without
sacrificing much sensitivity.
Simplicity in design may therefore have afforded a higher quality result by
reducing these delays and publishing its result years sooner,
and so implemented \cern's goal of
``accelerating science''~\cite{Radford:1551933}.

% this extra line activates enlarged spacing for the final paragraph lol
